A Dependently Typed Multi-Stage Calculus

### 2019-08-17 main.tex
Reflects all easy comments.

To rigorously study properties of such an extension, we develop \LMD, which
is an extension of Hanada and Igarashi's typed calculus \LTP with dependent
types, and prove its properties including preservation, confluence, strong
normalization for full reduction, and progress for staged reduction.
Motivated by code generators that generate code whose type depends on
a value from outside of the quotations, we argue the significance of cross-stage
persistence in dependently typed multi-stage programming and certain type
equivalence that are not directly derived from reduction rules.

>>>>

To rigorously study properties of such an extension, we develop \LMD, which
is an extension of Hanada and Igarashi's typed calculus \LTP with dependent
types, and prove its properties including preservation, confluence, strong
normalization for full reduction, and progress for staged reduction.
Motivated by code generators that generate code whose type depends on
a value from outside of the quotations, we argue the significance of cross-stage
persistence in dependently typed multi-stage programming and certain type
equivalences that are not directly derived from reduction rules.

### 2019-08-17 text/formal-defnition.tex
Reflects all easy comments.

The prefix operators $\TW_\alpha, \TB_\alpha, \TBL_\alpha$, and
$\%_\alpha$ connect tighter than the three forms $\tau\ M$, $M\ N$,
$M\ A$ of applications, which are left-associative; The binders $\Pi$,
$\forall$, and $\lambda$ extend as far to the right as possible.
Thus, $\F\alpha.\TW_{\alpha} (\Pi x:\I.\text{Vector}\ 5)$ is
interpreted as
$\F\alpha.(\TW_{\alpha} (\Pi x:\I.(\text{Vector}\ 5)))$; and
$\Lambda\alpha.\lambda x:\I.\TB_\alpha x\ y$ means
$\Lambda\alpha.(\lambda x:\I.(\TB_\alpha x)\ y)$.

@@

\subsubsection{Kind Wellformedness and Kinding.}

The rules for kind wellformedness and kinding is a straightforward
adaptation from \LLF and \LTP, except for the following rule for type-level CSP.
\begin{center}
  \infrule[K-Csp]{
    \G \V \tau :: * @ A
  }{
    \G \V \tau :: * @ A\alpha
  }
\end{center}
Unlike the term level, type-level CSP is implicit.

@@

% \TConv
As in standard dependent type systems, \TConv{} allows us to replace the type of a term with an equivalent one.
For example, assuming integers and arithmetic, a value of type $\textrm{Vector}\ (4+1)$ can also has type $\textrm{Vector}\ 5$ because of \TConv{}.

@@

% \QPercentの説明
The only rule that deserves elaboration is the last rule \QPercent.
Intuitively, it means that the CSP operator applied to term $M$ can be
removed if $M$ is also well typed at the next stage \(A\alpha\).
For example, constants do not depend on the stage (see \TConst) and
so \(\G\V \%_\alpha c \E c : \tau @ A\alpha\) holds but variables
do depend on stages and so this rule does not apply.

>>>>

The prefix operators $\TW_\alpha, \TB_\alpha, \TBL_\alpha$, and
$\%_\alpha$ connect tighter than the three forms $\tau\ M$, $M\ N$,
$M\ A$ of applications, which are left-associative. The binders $\Pi$,
$\forall$, and $\lambda$ extend as far to the right as possible.
Thus, $\F\alpha.\TW_{\alpha} (\Pi x:\I.\text{Vector}\ 5)$ is
interpreted as
$\F\alpha.(\TW_{\alpha} (\Pi x:\I.(\text{Vector}\ 5)))$; and
$\Lambda\alpha.\lambda x:\I.\TB_\alpha x\ y$ means
$\Lambda\alpha.(\lambda x:\I.(\TB_\alpha x)\ y)$.

@@

\subsubsection{Kind Well-formedness and Kinding.}

The rules for kind well-formedness and kinding is a straightforward
adaptation from \LLF and \LTP, except for the following rule for type-level CSP.
\begin{center}
  \infrule[K-Csp]{
    \G \V \tau :: * @ A
  }{
    \G \V \tau :: * @ A\alpha
  }
\end{center}
Unlike the term level, type-level CSP is implicit.

@@

% \TConv
As in standard dependent type systems, \TConv{} allows us to replace the type
of a term with an equivalent one. For example, assuming integers and
arithmetic, a value of type $\textrm{Vector}\ (4+1)$ can also have type
$\textrm{Vector}\ 5$ because of \TConv{}.

@@

% \QPercentの説明
The only rule that deserves elaboration is the last rule \QPercent.
Intuitively, it means that the CSP operator applied to term $M$ can be
removed if $M$ is also well-typed at the next stage \(A\alpha\).
For example, constants do not depend on the stage (see \TConst) and
so \(\G\V \%_\alpha c \E c : \tau @ A\alpha\) holds but variables
do depend on stages and so this rule does not apply.

### 2019-08-17 text/informal-overview.tex
Reflects all easy comments.

The subscript $\alpha$ in $\TB_\alpha$ and $\TBL_\alpha$ is a \textit{stage
  variable}\footnote{%
  In Hanada and Igarashi~\cite{Hanada2014}, it was called a
  \textit{transition variable}, which is derived from correspondence
  to modal logic, studied by Tsukada and Igarashi~\cite{Tsukada}.} and
a sequence of stage variables is called a \textit{stage}.  Intuitively,
they represent the depth of nested brackets.  Stage variables can be
abstracted by $\Lambda\alpha.M$ and instantiated by an application
$M\ A$ to stages.  For example,
$\Lambda\alpha.\TB_\alpha ((\lambda x:\I.x+10)\ 5)$ is a code value,
where \(\alpha\) is abstracted.  If it is applied to \(\alpha_1 \cdots \alpha_n\), \(\TB_\alpha\) becomes \(\TB_{\alpha_1} \cdots \TB_{\alpha_n}\); in particular,
if \(n = 0\), \(\TB_\alpha\) disappears.  So, an
application of $\Lambda\alpha.\TB_\alpha ((\lambda x:\I.x+10)\ 5)$
to the empty sequence \(\varepsilon\) reduces to
(unquoted) \((\lambda x:\I.x+10)\ 5\) and to 15.  In other words, application of of
\(\Lambda\)-abstraction to $\varepsilon$ corresponds to \texttt{run}.
% If \(\alpha\) is instantiated with, say, \(\beta\gamma\), then
% \(\TB_\alpha\) becomes nested brackets \(\TB_{\beta} \TB_\gamma\).
This is expressed by the following reduction rule:
\begin{align*}
	(\Lambda\alpha.M)\ A \longrightarrow M[\alpha\mapsto A]
\end{align*}
where stage substitution \([\alpha \mapsto A]\) manipulates the nesting of
\(\TB_\alpha\) and \(\TBL_\alpha\) (and also \(\%_\alpha\) as we see later).

@@

Second, an indexed type such as Vector\ 3 or Vector\ $x$ is well
formed only at the stage(s) where the index term is well typed.  Since
constant \(3\) is well typed at every stage (if it is declared in the
signature), Vector\ 3 is a well-formed type at every stage, too.
However, Vector\ $M$ is well formed only at the stage where index term
$M$ is typed.  Thus, the kinding judgment
of \LMD takes the form \(\G\V \tau :: K @ A\), where stage $A$ stands for
where \(\tau\) is well formed.  For example,
given \(\text{Vector}:: \I \rightarrow *\) in the signature \(\Sigma\),
\(x:\I@\varepsilon \V \text{Vector }x :: * @\varepsilon\) can be
derived but
neither
\(x:\I@\alpha \V \text{Vector }x :: * @\varepsilon\)
nor 
\(x:\I@\varepsilon \V \text{Vector }x :: * @\alpha\)
can be.

>>>>

The subscript $\alpha$ in $\TB_\alpha$ and $\TBL_\alpha$ is a \textit{stage
  variable}\footnote{%
  In Hanada and Igarashi~\cite{Hanada2014}, it was called a
  \textit{transition variable}, which is derived from correspondence
  to modal logic, studied by Tsukada and Igarashi~\cite{Tsukada}.} and
a sequence of stage variables is called a \textit{stage}.  Intuitively,
they represent the depth of nested brackets.  Stage variables can be
abstracted by $\Lambda\alpha.M$ and instantiated by an application
$M\ A$ to stages.  For example,
$\Lambda\alpha.\TB_\alpha ((\lambda x:\I.x+10)\ 5)$ is a code value,
where \(\alpha\) is abstracted.  If it is applied to \(\alpha_1 \cdots \alpha_n\), \(\TB_\alpha\) becomes \(\TB_{\alpha_1} \cdots \TB_{\alpha_n}\); in particular,
if \(n = 0\), \(\TB_\alpha\) disappears.  So, an
application of $\Lambda\alpha.\TB_\alpha ((\lambda x:\I.x+10)\ 5)$
to the empty sequence \(\varepsilon\) reduces to
(unquoted) \((\lambda x:\I.x+10)\ 5\) and to 15.  In other words, application of
\(\Lambda\)-abstraction to $\varepsilon$ corresponds to \texttt{run}.
% If \(\alpha\) is instantiated with, say, \(\beta\gamma\), then
% \(\TB_\alpha\) becomes nested brackets \(\TB_{\beta} \TB_\gamma\).
This is expressed by the following reduction rule:
\begin{align*}
	(\Lambda\alpha.M)\ A \longrightarrow M[\alpha\mapsto A]
\end{align*}
where stage substitution \([\alpha \mapsto A]\) manipulates the nesting of
\(\TB_\alpha\) and \(\TBL_\alpha\) (and also \(\%_\alpha\) as we see later).

@@

Second, an indexed type such as Vector\ 3 or Vector\ $x$ is well
formed only at the stage(s) where the index term is well-typed.  Since
constant \(3\) is well-typed at every stage (if it is declared in the
signature), Vector\ 3 is a well-formed type at every stage, too.
However, Vector\ $M$ is well-formed only at the stage where index term
$M$ is typed.  Thus, the kinding judgment
of \LMD takes the form \(\G\V \tau :: K @ A\), where stage $A$ stands for
where \(\tau\) is well-formed.  For example,
given \(\text{Vector}:: \I \rightarrow *\) in the signature \(\Sigma\),
\(x:\I@\varepsilon \V \text{Vector }x :: * @\varepsilon\) can be
derived but
neither
\(x:\I@\alpha \V \text{Vector }x :: * @\varepsilon\)
nor 
\(x:\I@\varepsilon \V \text{Vector }x :: * @\alpha\)
can be.

### 2019-08-17 text/introduction.tex
Reflects all easy comments.

MetaOCaml~\cite{calcagno2003implementing,oleg2014} is an extension of
OCaml\footnote{\url{http://ocaml.org}} with special constructs for
multi-stage programming, including brackets and escape, which are
(hygienic) quasi-quotation, and \texttt{run}, which is similar to
\texttt{eval} in Lisp, and cross-stage persistence
(CSP)~\cite{MetaML}.  Programmers can easily write code generators by
using these features.  Moreover, MetaOCaml is equipped with a powerful
type system for safe code generation and execution.  The notion of
code types is introduced to prevent code values that represent
ill-typed expressions from being generated.  For example, a quotation
of expression \texttt{1 + 1} is given type \texttt{int code}
and a code-generating function, which takes a code value \(c\) as an
argument and returns \(c \texttt{ + } c\), is given type \texttt{int
  code -> int code} so that it cannot be applied to, say, a quotation
of \texttt{"Hello"}, which is given type \texttt{string
  code}.  Ensuring safety for \verb|run| is more challenging because
code types per by themselves do not guarantee that the execution of code values never
results in unbound variable errors.  Taha and
Nielsen~\cite{taha2003environment} introduced the notion of
environment classifiers to address the problem, developed a type
system to ensure not only type-safe composition but also type-safe
execution of code values, and proved a type soundness theorem (for a formal calculus \(\lambda^\alpha\) modeling a pure subset of MetaOCaml).

@@

One natural idea to address this problem is the introduction of dependent types
to express the size of data structure in static types~\cite{Xi98}.
% Dependent types are types which are dependent on values.
% We can use dependent types for securer programming like Xi and Pfenning
For example, we could declare vector types indexed by the size of
vectors as follows.
\begin{verbatim}
    Vector :: Int -> *
\end{verbatim}
\verb|Vector| is a type constructor that takes an integer (which
represents the length of vectors): for example, \verb|Vector 3| is the
type for vectors whose lengths are 3.  Then, our hope is to specialize
vector/matrix functions with respect to their size and get a piece of
function code, whose type respects the given size, \emph{provided at
  specialization time}.  For example, we would like to specialize a
function to add two vectors with respect to the size of vectors, that
is, to implement a code generator that takes a (nonnegative) integer $n$ as an
input and generates a piece of function code of type
\verb|(Vector |$n$\verb| -> Vector |$n$\verb| -> Vector |$n$\verb|) code|.

@@

\subsection{Our Work}
In this paper, we develop a new multi-stage calculus \LMD by extending
the existing multi-stage calculus \LTP\cite{Hanada2014} with dependent
types and study its properties.  We base our work on \LTP, in which
the four multi-stage constructs are handled slightly differently from
MetaOCaml, because its type system and semantics are arguably simpler
than \(\lambda^\alpha\)~\cite{taha2003environment}, which formalizes
the design of MetaOCaml more faithfully.  Dependent types are based on
\LLF~\cite{attapl}, which has one of the simplest form of dependent
types.  Our technical contributions are summarized as follows:
\begin{itemize}
\item We give a formal definition of \LMD with its syntax, 
  type system and two kinds of reduction---full and staged reduction.
\item We show preservation, strong normalization, and confluence for
  full reduction; and show unique decomposition (and progress as its
  corollary) for staged reduction, which formalizes program execution.
\end{itemize}
The combination of multi-stage programming and dependent types has
been discussed by Brady and Hammond~\cite{brady2006dependently} but,
to our knowledge, our work is a first formal calculus of dependently
typed multi-stage programming.

>>>>

MetaOCaml~\cite{calcagno2003implementing,oleg2014} is an extension of
OCaml\footnote{\url{http://ocaml.org}} with special constructs for
multi-stage programming, including brackets and escape, which are
(hygienic) quasi-quotation, and \texttt{run}, which is similar to
\texttt{eval} in Lisp, and cross-stage persistence
(CSP)~\cite{MetaML}.  Programmers can easily write code generators by
using these features.  Moreover, MetaOCaml is equipped with a powerful
type system for safe code generation and execution.  The notion of
code types is introduced to prevent code values that represent
ill-typed expressions from being generated.  For example, a quotation
of expression \texttt{1 + 1} is given type \texttt{int code}
and a code-generating function, which takes a code value \(c\) as an
argument and returns \(c \texttt{ + } c\), is given type \texttt{int
  code -> int code} so that it cannot be applied to, say, a quotation
of \texttt{"Hello"}, which is given type \texttt{string
  code}.  Ensuring safety for \verb|run| is more challenging because
code types by themselves do not guarantee that the execution of code values never
results in unbound variable errors.  Taha and
Nielsen~\cite{taha2003environment} introduced the notion of
environment classifiers to address the problem, developed a type
system to ensure not only type-safe composition but also type-safe
execution of code values, and proved a type soundness theorem (for a formal calculus \(\lambda^\alpha\) modeling a pure subset of MetaOCaml).

@@

One natural idea to address this problem is the introduction of dependent types
to express the size of data structures in static types~\cite{Xi98}.
% Dependent types are types which are dependent on values.
% We can use dependent types for securer programming like Xi and Pfenning
For example, we could declare vector types indexed by the size of
vectors as follows.
\begin{verbatim}
    Vector :: Int -> *
\end{verbatim}
\verb|Vector| is a type constructor that takes an integer (which
represents the length of vectors): for example, \verb|Vector 3| is the
type for vectors whose lengths are 3.  Then, our hope is to specialize
vector/matrix functions with respect to their size and get a piece of
function code, whose type respects the given size, \emph{provided at
  specialization time}.  For example, we would like to specialize a
function to add two vectors with respect to the size of vectors, that
is, to implement a code generator that takes a (nonnegative) integer $n$ as an
input and generates a piece of function code of type
\verb|(Vector |$n$\verb| -> Vector |$n$\verb| -> Vector |$n$\verb|) code|.

@@

\subsection{Our Work}
In this paper, we develop a new multi-stage calculus \LMD by extending
the existing multi-stage calculus \LTP\cite{Hanada2014} with dependent
types and study its properties.  We base our work on \LTP, in which
the four multi-stage constructs are handled slightly differently from
MetaOCaml, because its type system and semantics are arguably simpler
than \(\lambda^\alpha\)~\cite{taha2003environment}, which formalizes
the design of MetaOCaml more faithfully.  Dependent types are based on
\LLF~\cite{attapl}, which has one of the simplest forms of dependent
types.  Our technical contributions are summarized as follows:
\begin{itemize}
\item We give a formal definition of \LMD with its syntax, 
  type system and two kinds of reduction---full and staged reduction.
\item We show preservation, strong normalization, and confluence for
  full reduction; and show unique decomposition (and progress as its
  corollary) for staged reduction, which formalizes program execution.
\end{itemize}
The combination of multi-stage programming and dependent types has
been discussed by Brady and Hammond~\cite{brady2006dependently} but,
to our knowledge, our work is a first formal calculus of dependently
typed multi-stage programming.

### 2019-08-17 text/properties.tex
Reflects all easy comments.

@@

\begin{proof}
  By simultaneous induction on derivations.
 \qed
\end{proof}

@@

The following theorem Unique Decomposition ensures that, every typed
term is either a value or can be uniquely decomposed to an evaluation
context and a redex, ensuring that a well typed term is not
immediately stuck and the staged semantics is deterministic.

@@

\begin{corollary}[Progress]
	If $\G$ does not have any variable declared at stage $\varepsilon$ and $\G \V M : \tau  @ A$, then
	$ M \in V^A $ or $M'$ exists such that $M \longrightarrow_s M'$.
\end{corollary}

>>>>

We proved the next two leammas by simultaneous induction on derivations.

@@

@@

The following theorem Unique Decomposition ensures that, every typed
term is either a value or can be uniquely decomposed to an evaluation
context and a redex, ensuring that a well-typed term is not
immediately stuck and the staged semantics is deterministic.

@@

\begin{corollary}[Progress]
	If $\G$ does not have any variable declared at stage $\varepsilon$ and $\G \V M : \tau  @ A$, then
	$ M \in V^A $ or there exists $M'$ such that $M \longrightarrow_s M'$.
\end{corollary}

### 2019-08-17 text/related-work.tex
Reflect comments from John.
Reflects all easy comments.

MetaOCaml is a programming language with quoting, unquoting, run, and
CSP.  Kiselyov~\cite{8384206} describes many applications of MetaOCaml, including filtering in signal processing, matrix-vector product,
and DSL compiler.

@@

There is much work on dependent types and most of them are affected by
the pioneering work by Martin-L\"{o}f~\cite{martin1973intuitionistic}.
Among many dependent type systems such as
$\lambda^\Pi$~\cite{Meyer1986}, The Calculus of
Constructions~\cite{coquand:inria-00076024}, and Edinburgh
LF~\cite{harper1993framework}, we base our work on \LLF~\cite{attapl}
(which is quite close to $\lambda^\Pi$ and Edinburgh LF) due to its
simplicity.  It has been well known that dependent types are useful to
express detailed properties of data structure at the type level such
as the size of data structures~\cite{Xi98} and typed abstract syntax
trees~\cite{DBLP:conf/dsl/LeijenM99,DBLP:conf/popl/XiCC03}.  The
vector addition discussed in Section~\ref{sec:formal} is also such an
example.

@@

Although there are studies on combinations of multi-stage programming and other programming features such as mutable cells~\cite{kiselyov2016refined},
control operators~\cite{KameyamaKiselyovShan09PEPM,oishi2017staging},
a combination with dependent types has been little studied.
One exception is Brady and Hammond~\cite{brady2006dependently},
who have discussed a
combination of dependently typed programming with staging in the
style of MetaOCaml to implement a staged interpreter, which is
statically guaranteed to generate well typed code.  However, they
focused on concrete programming examples and there is no theoretical investigation
of the programming language they used.

>>>>

MetaOCaml is a programming language with quoting, unquoting, run, and
CSP.  Kiselyov~\cite{8384206} describes many applications of MetaOCaml, including filtering in signal processing, matrix-vector product,
and a DSL compiler.

@@

There is much work on dependent types and most of it are affected by
the pioneering work by Martin-L\"{o}f~\cite{martin1973intuitionistic}.
Among many dependent type systems such as
$\lambda^\Pi$~\cite{Meyer1986}, The Calculus of
Constructions~\cite{coquand:inria-00076024}, and Edinburgh
LF~\cite{harper1993framework}, we base our work on \LLF~\cite{attapl}
(which is quite close to $\lambda^\Pi$ and Edinburgh LF) due to its
simplicity.  It is well known that dependent types are useful to
express detailed properties of data structures at the type level such
as the size of data structures~\cite{Xi98} and typed abstract syntax
trees~\cite{DBLP:conf/dsl/LeijenM99,DBLP:conf/popl/XiCC03}.  The
vector addition discussed in Section~\ref{sec:formal} is also such an
example.

@@

Although there are studies on combinations of multi-stage programming and
other programming features such as mutable cells~\cite{kiselyov2016refined} and
control operators~\cite{KameyamaKiselyovShan09PEPM,oishi2017staging}, a
combination with dependent types has been little studied. One exception is
Brady and Hammond~\cite{brady2006dependently}, who have discussed a
combination of dependently typed programming with staging in the style of
MetaOCaml to implement a staged interpreter, which is statically guaranteed
to generate well-typed code. However, they focused on concrete programming
examples and there is no theoretical investigation of the programming
language they used.
